---
title: "Templated analysis"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---


```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(
              fig.width=2*fig.dim,
              fig.height=fig.dim,
              fig.align='center',
              error=FALSE )
library(Matrix)
library(raster)
library(rgeos)
library(landsim)
# library(parallel)
# this.lapply <- function (...) { mclapply(...,mc.cores=detectCores() ) }
# need to check for available memory, working in parallel.
this.lapply <- lapply
```

Command-line for this file:
```{r call, echo=FALSE}
paste(commandArgs(),collapse=" ")
```

```{r analysis_setup, echo=FALSE, cache=TRUE, cache.extra=sourced.files}
required <- c("pop","demog")
usage <- " \
Before rendering this file, the following must be defined:
    pop : a population object \
    demog : a demography object \
but they can't be found.
"
if ( !all(sapply(required,exists)) ) { stop(usage) }
demog <- setup_demography( demog, pop )
```

# Set-up:

Here are the descriptions of the population and the demography:

- **population**: `r (pop$description)`
- **demography**: `r (demog$description)`

Here is the habitat:
```{r show_habitat, fig.width=3*fig.dim}
layout(t(1:3))
plot(pop$habitat,main="habitat")
thab <- pop$habitat
values(thab) <- ifelse( pop$accessible, 1, NA )
plot(thab,main="accessible")
thab <- pop$habitat
values(thab) <- ifelse( pop$habitable, 1, NA )
plot(thab,main="habitable")
```

```{r plfun, include=FALSE}
# helper plotting function
pl <- function (x,zlim=range(0,x,finite=TRUE),...) {
    hab <- do.call(stack,list(pop$habitat)[rep(1,NCOL(x))]); 
    values(hab)[pop$habitable] <- x; 
    plot(hab,nr=1,zlim=zlim,legend.width=2,legend.mar=12,...) 
}
```


## Stability check

First, let's see if this is stable.
Here's the mean growth rate per individual at low density:
```{r stability_check, fig.width=3*fig.dim}
invasion.rate <- intrinsic_growth( pop, demog )
pl(invasion.rate-1)
```

# Statistics of the wave

We'll record a few statistics about the wave of advantageous alleles as it spreads,
including its *radius* and *volume*.
(We can get surface area by looking at changing volume or the area occupied by heterozygotes.)
Here's the set-up for computing these,
which will be functions of $N$ passed as summaries.
```{r geom_stats}
# distance to the mutation center for each habitable cell
volume_fun <- function (N) { colSums(N>0) }
# will compute 'dists' below
radius_fun <- function (N) { apply( N, 2, function (x) { suppressWarnings( max( dists[ x>0 ] ) ) } ) }
```


# Wave of adaptation


Choose a location for a single mutation, and make sure it establishes:
```{r init_muts, fig.width=fig.dim, cache=TRUE, dependson="analysis_setup"}
mutloc <- list( cell.number=sample.int(nhabitable(pop),1) )
mutloc$center <- SpatialPoints(xyFromCell(pop$habitat,which(pop$habitable)[mutloc$cell.number]))
dists <- pointDistance( mutloc$center, xyFromCell(pop$habitat,which(pop$habitable)), lonlat=FALSE )
for (ntries in 1:100) {
    pop <- set_N( pop, i=which(pop$habitable), j="aA", 
             value=ifelse(1:nrow(pop$N)==mutloc$cell.number,1,0) )
    sim <- simulate_pop( pop, demog, times=seq(0,100,length.out=11),
                     summaries= list( 
                             totals=function(N){colSums(N)},
                             volume=volume_fun,
                             radius=radius_fun
                         ),
                     stop.fun=function(N){ sum(N)==0 } )
    total.aA <- sim$summaries[[1]][nrow(sim$summaries[[1]]),"aA"]
    if ( total.aA>0 ) { break }
}
if (total.aA==0) { stop(paste("Failed to establish after",ntries,"tries.")) }
```

Now, run longer:
```{r run_longer, cache=TRUE, dependson="init_muts"}
sim <- extend_simulation( sim, pop, demog, times=seq(sim$t,1000,length.out=91)[-1],
                 summaries= list( 
                             totals=function(N){ colSums(N) },
                             volume=volume_fun,
                             radius=radius_fun
                         ),
                 stop.fun= function(N){sum(N)==0} )
```
```{r plot_total_numbers}
suppressWarnings( matplot( sim$summary.times, sim$summaries[["totals"]], type='l', lty=1, log='y', ylab="total numbers" ) )
legend("topright",lty=1,col=1:3,legend=pop$genotypes)
```
```{r plot_run_longer, cache=TRUE, dependson="run_longer", fig.show="animate", aniopts="controls", interval=0.1, fig.width=3*fig.dim}
plot(sim,pop)
```

## Speed of expansion

To get the speed of expansion
we'll compare the spread to the minimum enclosing circle,
for heterozygotes and for homozygotes.
```{r expansion}
# plot:
matplot( sim$summary.times, sim$summaries[["radius"]][,c("aA","AA")], type='l', lty=1, col=seq_along(pop$genotypes)[-1],
        xlab="generation", main='radius', ylab='distance (m)')
legend( "bottomright", legend=pop$genotypes[-1], lty=1, col=seq_along(pop$genotypes)[-1])
AA.lims <- c(1.2,0.8)*range(sim$summaries[["radius"]][,"AA"],finite=TRUE)
AA.usethese <- which( ( sim$summaries[["radius"]][,"AA"] > AA.lims[1] ) 
                      & ( sim$summaries[["radius"]][,"AA"] < AA.lims[2] ) )
if (length(AA.usethese)>1) {
    AA.speed <- diff( sim$summaries[["radius"]][range(AA.usethese),"AA"] ) / diff( sim$summary.times[range(AA.usethese)] )
    points(  sim$summary.times[range(AA.usethese)], sim$summaries[["radius"]][range(AA.usethese),"AA"], pch=20, col='red' )
    abline( sim$summaries[["radius"]][min(AA.usethese),"AA"]-AA.speed*sim$summary.times[min(AA.usethese)], AA.speed, col='red', lty=3 )
    AA.speed
} else {
    AA.speed <- NA
}
```

## Radius-volume relationship

Perhaps a useful measure of the geometry of the habitat
is how much habitat has been covered by the wave as a function of its radius?
Here, "volume" is the total number of occupied cells;
this should be similar to the total numbers of individuals reported above.
```{r volume}
matplot( sim$summary.times, sim$summaries[["volume"]][,c("aA","AA")], type='l', lty=1, col=seq_along(pop$genotypes),
        xlab="generation", main='volume', ylab='volume (m^2)')
legend( "bottomright", legend=pop$genotypes[-1], lty=1, col=seq_along(pop$genotypes))
geoms <- data.frame( 
                    time=sim$summary.times, 
                    het.total=sim$summaries[["totals"]][,"aA"], 
                    het.radius=sim$summaries[["radius"]][,"aA"], 
                    het.volume=sim$summaries[["volume"]][,"aA"],
                    hom.total=sim$summaries[["totals"]][,"AA"], 
                    hom.radius=sim$summaries[["radius"]][,"AA"], 
                    hom.volume=sim$summaries[["volume"]][,"AA"]
                )
with( geoms, plot( het.radius, het.total, 
     pch=20, cex=0.5, xlab='radius (m)', ylab='volume (m^2)' ) )
```


## Width of the wave

To compute the width of the wave,
here is the difference between the furthest occurence that *A* is at 25% and the furthest location it is at 75%:
```{r wave_width}
A.freqs <- (sim$N[,2,]/2 + sim$N[,3,])/(sim$N[,1,] + sim$N[,2,] + sim$N[,3,])
quant.dists <- cbind(
                     max = apply( A.freqs, 2, function (x) { if (any(x[!is.na(x)]>0.5)) { max( dists[x>0], na.rm=TRUE ) - max( dists[x>0.5], na.rm=TRUE ) } else { NA } } ), 
                     iqr = apply( A.freqs, 2, function (x) { if (any(x[!is.na(x)]>0.75)) { max( dists[x>0.25], na.rm=TRUE ) - max( dists[x>0.75], na.rm=TRUE ) } else { NA } } )
                 )
width.usethese <- which( ( sim$times >= min(sim$summary.times[AA.usethese]) ) & ( sim$times <= max(sim$summary.times[AA.usethese]) ) )
est.widths <- colMeans(quant.dists[width.usethese,,drop=FALSE],na.rm=TRUE)
```


## Comparison to theory

**Generation time:**
We also need the generation time,
because what we're calling a generation isn't, really.
Generation time in this simulation is determined by the probability of death per generation,
which here is `r demog$prob.survival`.

**Dispersal distance:**
Simulation doesn't match up with theory so well above.
But, we're using a naive estimate of the dispersal distance, $\sigma$.
We can get a better estimate from the migration matrix itself;
```{r empirical_sigma,fig.width=2*fig.dim}
seed.sigmas <- dispersal_distance(demog$seed.migration$M,pop)
pollen.sigmas <- dispersal_distance(demog$pollen.migration$M,pop)
mean.sigma <- (mean(seed.sigmas)+mean(pollen.sigmas))/2
c(empirical=mean.sigma,naive=(demog$pollen.migration$sigma+demog$seed.migration$sigma)/2)
layout(t(1:2))
pl(seed.sigmas,main="sigma for seeds")
pl(pollen.sigmas,main="sigma for pollen")
# this will have length zero if it doesn't exist
s <- (demog$prob.germination$s-1)
ctab <- data.frame(
           speed = c( expected = 2 * mean.sigma * sqrt(s) * (1-demog$prob.survival) ,
                      observed = AA.speed ),
           width = c( expected = 2 * mean.sigma / sqrt(s) ,
                      observed = est.widths[2] ) )
floor(ctab)
```


# Probability of Establishment

## Sample mutation locations


Sample and initiate mutations:
```{r sample_muts, fig.width=fig.dim, cache=TRUE, dependson="analysis_setup"}
nhood.radius <- 1000
many.sims <- this.lapply( 1:36, function (dummy) {
        nhoods <- sample_neighborhoods( x=mask(habitat,habitat,maskvalue=0), n=60, 
                               radius=nhood.radius, separation=2*nhood.radius )
        pop <- set_N( pop, i=which(pop$habitable), j="aA", 
                 value=ifelse(which(pop$habitable)%in%nhoods$center.cells,1,0) )
        sim <- simulate_pop( pop, demog, times=c(1,50,100),
                         summaries=c( census_neighborhoods( pop, nhoods ),  
                                    list( totals=function(N){colSums(N)} ) ) )
        return( list( centers=nhoods$centers, summaries=sim$summaries ) )
     } )
sim.results <- list( centers = SpatialPoints( do.call(rbind,lapply(lapply(many.sims,"[[","centers"),coordinates)) ),
                    established = do.call( c, lapply( many.sims, function (ms)
                       {( (ms$summaries[["aA"]]+ms$summaries[["AA"]])[nrow(ms$summaries[["aA"]]),] > 0 )} ) )
                    )
```

As a result, we have a set of initial locations
and a binary variable (whether or not they established).
```{r interp_prob, fig.height=1.5*fig.dim, fig.width=1.5*fig.dim}
plot( pop$habitat )
points(sim.results$centers,pch=20,col=1+sim.results$established)
```

The overall probability of establishment is `r mean(sim.results$established)`.



# Raw data:

Here are the coordinates of mutations, along with whether they established, as a csv:
```{r results}
write.csv( cbind( coordinates(sim.results$centers), sim.results$established ), file="", row.names=FALSE )
```

```{r show_plfun}
pl
```
