---
title: "Disconnected habitats"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---

```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(
              fig.width=2*fig.dim,
              fig.height=fig.dim,
              fig.align='center',
              error=FALSE )
library(Matrix)
library(raster)
library(rgeos)
library(landsim)
```

Consider the **saline** habitat in the central Mojave:
```{r setup_demog, cache=TRUE}
source("habitats/central_mojave_saline.R",chdir=TRUE)
source("populations/high-density.R",chdir=TRUE)
source("demographies/long-distance-gaussian_neutral.R")
demog <- setup_demography(demog,pop)
```

This habitat is broken up into pieces.

One iteration of the migration matrix
=====================================

Suppose we only used one iteration of the migration matrix.
How much migration is there between the disconnected pieces?
```{r clump_habitat, fig.height=2*fig.dim, fig.width=2*fig.dim, cache=TRUE, depends="setup_demog"}
habclumps <- clump(pop$habitat)
# matrix of pairwise total migration rates between clumps
clumpmig <- matrix( nrow=length(unique(values(habclumps)[pop$habitable])), ncol=length(unique(values(habclumps)[pop$habitable])) )
for (k in 1:nrow(clumpmig)) {
    kk <- which( values(habclumps) == k )
    for (j in 1:nrow(clumpmig)) {
        jj <- which( values(habclumps) == j )
        clumpmig[j,k] <- sum( demog$pollen.migration$M[jj,kk] )
    }
}
# use '%&%' to find equivalency classes of connecteds
clump.adj <- (clumpmig>0)
diag(clump.adj) <- TRUE
clump.clump <- clump.adj
for (k in 1:10) {
    clump.clump <-  clump.clump %&% clump.clump
}
# here are the clusters of communicating elements
clump.clusters <- unique(apply(clump.clump,1,which))
cluster.num <- numeric(nrow(clumpmig))
cluster.num[ unlist(clump.clusters) ] <- rep(seq_along(clump.clusters),sapply(clump.clusters,length))
habclusters <- habclumps
values(habclusters) <- cluster.num[values(habclumps)]
# number of clumps per cluster
table(table(cluster.num))
cluster.cols <- sample(rainbow(length(clump.clusters)))
plot(habclusters,col=cluster.cols)
```

How far are the closest ones apart?
Note these are distances between *representative* points, not the shortest distances.
```{r clump_dists, fig.height=2*fig.dim, fig.width=2*fig.dim, cache=TRUE, depends="clump_habitat"}
# matrix of pairwise distances between clumps
refcells <- tapply(1:ncell(habclumps),values(habclumps),sample,size=1)
refpoints <- xyFromCell( pop$habitat, refcells )
clumpdist <- pointDistance( refpoints, lonlat=FALSE, allpairs=TRUE )
```

Let's see which the clumps would be if we allowed migration up to a certain distance.
```{r clump_sizes, fig.height=2*fig.dim, fig.width=2*fig.dim}
clump_cutoff <- function (maxdist,clumpdist,mindist=0,...) {
    adj <- as( ( is.na(clumpdist) | ( ( clumpdist <= maxdist ) & ( clumpdist >= mindist ) ) ), "ngCMatrix" )
    for (k in 1:10) { adj <- adj %&% adj }
    clump.clusters <- unique(apply(adj,1,which))
    cluster.num <- numeric(nrow(clumpdist))
    cluster.num[ unlist(clump.clusters) ] <- rep( seq_along(clump.clusters),
                                                 sapply(clump.clusters,length) )
    habclusters <- habclumps
    values(habclusters) <- cluster.num[values(habclumps)]
    cluster.cols <- sample(rainbow(length(clump.clusters)))
    plot(habclusters,col=cluster.cols,...)
    return(habclusters)
}
clump_cutoff(5000,clumpdist); points(refpoints)
clump_cutoff(15000,clumpdist); points(refpoints)
clump_cutoff(25000,clumpdist); points(refpoints)
clump_cutoff(37000,clumpdist); points(refpoints)
```


Many iterations of the migration matrix
=====================================

In fact, `pollen.migration` applies the migration matrix `r length(demog$pollen.migration$n.weights)` times.
How much migration is there now?
We'll find this out by putting mass 1 on each clump, migrating it forward by one step,
and seeing where mass ends up.
```{r clump_habitat, fig.height=2*fig.dim, fig.width=2*fig.dim, cache=TRUE, depends="setup_demog"}
clumpmig2 <- sapply( unique(values(habclumps)[pop$habitable]), function (k) {
        x <- ifelse( values(habclumps)[pop$habitable] == k, 1, 0 )
        x <- x/sum(x)
        y <- migrate( x, demog$pollen.migration )
        tapply( y, values(habclumps)[pop$habitable], sum )
      } )
# use '%&%' to find equivalency classes of connecteds
clump2.adj <- (clumpmig2>0)
diag(clump2.adj) <- TRUE
clump2.clump <- clump2.adj
for (k in 1:10) {
    clump2.clump <-  clump2.clump %&% clump2.clump
}
# here are the clusters of communicating elements
clump2.clusters <- unique(lapply(1:nrow(clump2.clump), function (k) { which(clump2.clump[k,]) } ))
cluster.num2 <- numeric(nrow(clumpmig2))
cluster.num2[ unlist(clump2.clusters) ] <- rep(seq_along(clump2.clusters),sapply(clump2.clusters,length))
habclusters2 <- habclumps
values(habclusters2) <- cluster.num[values(habclumps)]
# number of clumps per cluster
table(table(cluster.num2))
cluster.cols2 <- sample(rainbow(length(unique(clump2.clusters))))
plot(habclusters2,col=cluster.cols2)
```

Let's look at how they break up when splitting up along different migration rates:
```{r clump_probs, fig.height=2*fig.dim, fig.width=2*fig.dim}
clump_cutoff(maxdist=Inf,mindist=0.0001,clumpmig2,main="connected by 1/10,000 migration")
clump_cutoff(maxdist=Inf,mindist=1e-6,clumpmig2,main="connected by 1/1,000,000 migration")
clump_cutoff(maxdist=Inf,mindist=1e-12,clumpmig2,main="connected by 1e-12 migration")
```
