---
title: "Local rate of drift"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---

```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(
              fig.width=2*fig.dim,
              fig.height=fig.dim,
              fig.align='center',
              error=FALSE )
library(Matrix)
library(raster)
library(rgeos)
library(landsim)
```

Goal
====

The goal is to measure the rate of local genetic drift
by the variance in allele frequency changes.

To do this, we'll pick some points in the landscape
and look at allele frequencies in circles of various radii
about those locations.


Simulation
==========

Here's the usual habitat set-up:
```{r setup_demog, cache=TRUE}
source("habitats/random_landscape.R")
source("demographies/gaussian-expsqrt.R")
# make it neutral!
demog$prob.germination$s <- 1
demog <- setup_demography(demog,pop)
```
Now, pick locations to census in:
```{r pick_centers, fig.width=fig.dim, cache=TRUE, dependson="setup_demog"}
nhoods <- sample_neighborhoods( pop$habitat, n=300, radius=300 )
radii <- c(1,3,9)*100
nhood.list <- lapply(radii, function (r) {
                     nhoods$neighborhoods <- make_circles(nhoods$centers, radii=r, proj4string=CRS(proj4string(pop$habitat)) )
                     return(nhoods)
              } )
census.funs <- lapply( nhood.list, census_neighborhoods, pop=pop )
names(census.funs) <- paste("r",radii,sep='')
plot(pop$habitat)
for (k in seq_along(radii)) {
    lines( nhood.list[[k]]$neighborhoods, col=k )
}
```
and run the simulation:
```{r run_sim, cache=TRUE, dependson="pick_centers"}
pop$N[] <- rpois( length(pop$N), outer(pop$N[,1],c(1/4,1/2,1/4),"*") )
sim <- simulate_pop( pop, demog, times=seq(0,1000,length.out=201),
                 summaries= c( unlist(census.funs), 
                          list( totals=colSums,
                                volume=function(N){colSums(N>0)}
                     ) ),
                 stop.fun=function(N){ sum(N)==0 } )
```


Allele frequencies
==================

Here are the overall genotype frequencies:
```{r total_freqs}
matplot(sim$summary.times, sim$summaries$totals/rowSums(sim$summaries$totals), type='l', lty=1, ylab="totals", xlab='time' )
legend("topright",lty=1,col=1:3,legend=pop$genotypes)
```

Now, here they are in each circle, for each radius:
```{r local_freqs}
freqs <- lapply( seq_along(radii), function (k) {
        aa <- sim$summaries[[paste("r",radii[k],".aa",sep='')]]
        aA <- sim$summaries[[paste("r",radii[k],".aA",sep='')]]
        AA <- sim$summaries[[paste("r",radii[k],".AA",sep='')]]
        return( (aA/2 + AA)/(aa+aA+AA) )
    } )
totals <- lapply( seq_along(radii), function (k) {
        aa <- sim$summaries[[paste("r",radii[k],".aa",sep='')]]
        aA <- sim$summaries[[paste("r",radii[k],".aA",sep='')]]
        AA <- sim$summaries[[paste("r",radii[k],".AA",sep='')]]
        return( 2*(aa + aA + AA) )
    } )
for (k in seq_along(radii)) {
    matplot( sim$summary.times, freqs[[k]], type='l', lty=1, ylim=c(0,1), 
            main=sprintf("r=%0.0f",radii[k]), xlab='time', ylab='frequency of A' )
}
```

Variance in allele frequencies
==============================

For each radius, we'll look at the variance of the allele frequency difference, 
divided by $p(1-p)$, where $p$ is the midpoint frequency.
Here, we should see no structure:
```{r freq_diffs}
fdiffs <- lapply( freqs, diff )
fmids <- lapply( seq_along(radii), function (k) {
       ( freqs[[k]][-1,] + freqs[[k]][-nrow(freqs[[k]]),] )/2
    } )
layout(t(seq_along(radii)))
for (k in seq_along(radii)) {
    plot( as.vector(fmids[[k]]), as.vector(fdiffs[[k]]), pch=20, xlab='freq', ylab='freq diff' )
}
```

In this plot, circle radius is proportional to the mean of $\sqrt{p_t(1-p_t)/\var{dp_t}}$:
```{r variances, fig.height=4*fig.dim, fig.width=4*fig.dim}
fvars <- lapply( seq_along(radii), function (k) {
            cov( fdiffs[[k]]/(fmids[[k]]*(1-fmids[[k]])) )
    } )
names(fvars) <- sprintf("r=%0.0f",radii)
plot(pop$habitat)
points(nhoods$centers,pch=20)
for (k in seq_along(radii)) {
    points( nhoods$centers, col=k, cex=radii[k]*sqrt(diag(fvars[[k]]))/2 )
}
```

This is expected to be close to the population size,
so to be comparable across ranges, we could scale this by census sizes in each circle.
Here's how well these match:
```{r var_vs_N}
svars <- sapply( seq_along(radii), function (k) {
            1/( diag(fvars[[k]]) * colMeans(totals[[k]]) )
    } )
matplot( sapply( totals, colMeans ), 1/sapply(fvars,diag), pch=1, log='xy',
        xlab='local abundance', ylab='local Neff') 
abline(0,1)
```
And here, the circles have radii proportional to $\sqrt{p_t(1-p_t)/N_t\var{dp_t}}$:
```{r scaled_var}
plot(pop$habitat)
points(nhoods$centers,pch=20)
for (k in seq_along(radii)) {
    points( nhoods$centers, col=k, cex=3*svars[,k] )
}
```

The mean value of $p_t(1-p_t)/N_t\var{dp_t}$ is as follows:
```{r mean_var_ratios}
data.frame( radius=radii,
           ratio = colMeans(  1/(sapply(fvars,diag)*sapply( totals, colMeans )) )
       )
```

# Raw data

Here's the raw numbers, in units of local effective popluation size:
```{r raw_Nes}
1/(sapply(fvars,diag))
```
