---
title: "Estimating the probability of establishment."
author: "Peter Ralph"
date: "`r Sys.Date()`"
---



```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(rgeos)
# library(gstat)
library(parallel)
devtools::load_all("../landsim")
```

```{r setup}
layerdir <- "../layers/Mojavset/Spatial/targets/biodiversity/binary"
layername <- "Eschscholzia_minutiflora_ssp_twisselmannii_broad_extent_avg.tif"
# habitats are listed with 0=absence, 1=presence.
habitat <- raster( file.path(layerdir,layername) )
```

Here's the raster:
```{r show_habitat, fig.width=fig.dim}
cropbox <- extent( c(
            xmin        = -2093993,
            xmax        = -1849272,
            ymin        =  1468496,
            ymax        =  1663602 ))
plot(habitat,main=layername)
# lines( cropbox )
habitat <- crop( habitat, cropbox )
```

To estimate the probability of establishment,
we'll: 

1.  Select some regularly-spaced 'mutant' locations,
2.  Initiate a simulation with single copies of advantageous alleles there, and
3.  Run it forwards to count how many establish.


# Demography

First we need a demographic set-up.
```{r demog_setup}
carrying.capacity <- 10
pop <- population( 
                  habitat = habitat,
                  accessible = !is.na(values(habitat)),
                  habitable = (!is.na(values(habitat)) & values(habitat)>0),
                  genotypes = c("aa","aA","AA"),
                  N = cbind( aa=rpois_raster(carrying.capacity*mask(habitat,habitat,maskvalue=0),only.values=TRUE),
                             aA=0, 
                             AA=0 )
             )

germination.args <- list( 
             r0 = 0.01,  # one in ten seeds will germinate at low densities
             s = 1.5,    # multiplicative selective benefit of the A allele
             carrying.capacity = carrying.capacity, 
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1,
                                     do.M=TRUE,
                                     population=pop
                                 )
         )
germination_fun <- function (N, ...) {
    out <- germination.args$r0 / ( 1 + migrate(germination.args$competition,x=rowSums(N))/germination.args$carrying.capacity )
    return( cbind( aa=out, aA=germination.args$s*out, AA=germination.args$s^2*out ) )
}

demog <- demography(
        prob.seed = 0.2,
        fecundity = 100,
        prob.germination = germination_fun,
        prob.survival = 0.9,
        pollen.migration = migration(
                            kern = function (x) { exp(-sqrt(x)) },
                            sigma = 300,
                            radius = 1200,
                            normalize = NULL,
                            do.M=TRUE,
                            population=pop
                     ),
        seed.migration = migration(
                            kern = "gaussian",
                            sigma = 100,
                            radius = 1200,
                            normalize = 1,
                            do.M=TRUE,
                            population=pop
                     ),
        genotypes = c("aa","aA","AA"),
        mating = mating_tensor( c("aa","aA","AA") )
    )
```


# Sample mutation locations

We choose locations by picking 200 random locations in the habitat,
and picking a maximal set of these so that their surrounding (circular) neighborhoods
do not overlap.
The *A* allele has advantage `r germination.args$s-1`,
so if it has not died out, after 10 generations, it should be present in about `r floor(germination.args$s^10)` copies
(excluding the effects of population structure).
The mean migration distance is about `r floor((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2)` meters,
so in that time it will have spread by at most `r 10*floor((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2*sqrt(germination.args$s))` meters.
We'll set the radii of the neighborhoods to be that value.
```{r sample_muts, fig.width=fig.dim}
nhood.radius <- 10*((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2*sqrt(germination.args$s))
nhoods <- sample_neighborhoods( x=mask(habitat,habitat,maskvalue=0), n=60, 
                       radius=nhood.radius, separation=2*nhood.radius )
plot(habitat)
points(nhoods$centers,pch="*",cex=2)
lines(nhoods$neighborhoods)
```

# Initiate mutations

First, we'll set each location to have one mutation.
```{r set_muts, fig.width=1.5*fig.dim, fig.height=1.5*fig.dim}
pop <- set_N( pop, i=which(pop$habitable), j="aA", 
         value=ifelse(which(pop$habitable)%in%nhoods$center.cells,1,0) )
colSums(pop$N)
# check this assigned correctly
nonz <- xyFromCell(pop$habitat,which(pop$habitable)[pop$N[,"aA"]>0],sp=TRUE)
plot(pop$habitat)
points(nonz,pch=20,col='red')
lines(nhoods$neighborhoods)
```

*Note:* an alternate strategy would be to extract out each neighborhood,
and simulate on these in isolation.
This might be quicker, but runs into difficulty when thinking about how to deal with nonlocal effects,
like competition between individuals in nearby cells.


# Simulate forwards

Now we need to run it forwards some generations,
and record the numbers of *A* alleles in each neighborhood.
```{r sim_muts}
plot.times <- seq(0,100,length.out=21)
sim <- simulate_pop( pop, demog, times=plot.times, 
                 summaries=c( census_neighborhoods( pop, nhoods ),  
                            list( totals=function(N){colSums(N)} ) ) )
```

Here's the numbers of *aA* and *AA* genotypes in each neighborhood.
We want to see good separation between those that have died out and those that haven't.
```{r show_mut_growth}
layout(t(1:2))
matplot(1+sim$summaries$aA,type='l',lty=1,main="aA counts per neighborhood",log='y')
matplot(1+sim$summaries$AA,type='l',lty=2,main="AA counts per neighborhood",log='y')
```

Here's the final result, after `r sim$t` generations,
with 
```{r show_muts, fig.width=1.5*fig.dim, fig.height=1.5*fig.dim}
nonz <- xyFromCell(pop$habitat,which(pop$habitable)[sim$N[,"aA",length(plot.times)]+sim$N[,"AA",length(plot.times)]>0],sp=TRUE)
established <- ( (sim$summaries[["aA"]]+sim$summaries[["AA"]])[nrow(sim$summaries[["aA"]]),] > 0 )
plot(pop$habitat)
points(nonz,pch=20,col='red',cex=0.5)
lines( nhoods$neighborhoods )
plot( nhoods$neighborhoods[established], add=TRUE,
     col=adjustcolor("red",0.25) )
```

# Do this many times

To get an estimate of the probabilty of establishment by location,
we need to do this many times, in many locations.
```{r many_muts}
many.sims <- mclapply( 1:36, function (dummy) {
        nhoods <- sample_neighborhoods( x=mask(habitat,habitat,maskvalue=0), n=60, 
                               radius=nhood.radius, separation=2*nhood.radius )
        pop <- set_N( pop, i=which(pop$habitable), j="aA", 
                 value=ifelse(which(pop$habitable)%in%nhoods$center.cells,1,0) )
        sim <- simulate_pop( pop, demog, times=c(1,50,100),
                         summaries=c( census_neighborhoods( pop, nhoods ),  
                                    list( totals=function(N){colSums(N)} ) ) )
        return( list( centers=nhoods$centers, summaries=sim$summaries ) )
     }, mc.cores=detectCores() )
sim.results <- list( centers = SpatialPoints( do.call(rbind,lapply(lapply(many.sims,"[[","centers"),coordinates)) ),
                    established = do.call( c, lapply( many.sims, function (ms)
                       {( (ms$summaries[["aA"]]+ms$summaries[["AA"]])[nrow(ms$summaries[["aA"]]),] > 0 )} ) )
                    )
```

As a result, we have a set of initial locations
and a binary variable (whether or not they established).
```{r interp_prob, fig.height=1.5*fig.dim, fig.width=1.5*fig.dim}
plot( pop$habitat )
points(sim.results$centers,pch=20,col=1+sim.results$established)
```

The overall probability of establishment is `r mean(sim.results$established)`.

Here's the raw results:
```{r results}
jsonlite::toJSON( cbind( coordinates(sim.results$centers), sim.results$established ) )
```
