---
title: "Probability of establishment"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(
              fig.width=3*fig.dim,
              fig.height=fig.dim,
              fig.align='center',
              error=FALSE )
library(Matrix)
library(raster)
library(rgeos)
library(landsim)
```

Command-line for this file:
```{r call, echo=FALSE}
paste(commandArgs(),collapse=" ")
```

```{r analysis_setup, echo=FALSE}
required <- c("habitat")
usage <- " \
Before rendering this file, the following must be defined: \
    habitat : a RasterLayer \
but they can't be found.
"
if ( !all(sapply(required,exists)) ) { stop(usage) }
```

Here's the standard demography,
motivated by Joshua tree,
and *modified to lower density*.
To change the selection coefficient we change `demog$prob.germination$s`.
The mean number of offspring at density $N$ offsets death when
$$
p_\text{seed} \times f \times \frac{ r_0 }{ 1 + N/C } = 1-p_\text{surv}
$$
which occurs when
$$
\left( \frac{ p_\text{seed} f r_0 }{1-p_\text{surv}} - 1 \right) = N/C
$$
Here we have
$$
\left( \frac{ 0.2 \times 100 \times .01 }{1-0.9} - 1 \right) = N/C = 1
$$
so we can make lower densities by reducing $C$, the carrying capacity.
```{r setup_demog, echo=FALSE}
demog <- demography(
        prob.seed = 0.2, fecundity = 100,
        prob.germination = vital( 
                 function (N, ...) {
                     out <- r0 / ( 1 + rowSums(N)/carrying.capacity )
                     return( cbind( aa=out, aA=(1+s)*out, AA=(1+s)^2*out ) )
                 },
                 r0 = 0.01,  s = 0.5 ),
        prob.survival = 0.9,
        pollen.migration = migration(
                 kern = function (x) { exp(-sqrt(x)) },
                 sigma = 300, radius = 1200, normalize = NULL ),
        seed.migration = migration(
                 kern = "gaussian", sigma = 100,
                 radius = 1200, normalize = 1 ),
        genotypes = c("aa","aA","AA")
    )
```

Here's the standard population, ditto:
```{r init_demog}
base.carrying.capacity <- 3.0 * prod(res(habitat))/270^2  # 12x lower than original, hovers around 1 per cell
habitable <- (!is.na(values(habitat)) & values(habitat)>0)
pop <- population( 
                  habitat = habitat,
                  # accessible = !is.na(values(habitat)),
                  accessible = habitable,
                  habitable = habitable,
                  genotypes = c("aa","aA","AA"),
                  carrying.capacity = base.carrying.capacity * values(habitat)[habitable],
                  N = cbind( aa=rpois(sum(habitable),base.carrying.capacity),
                             aA=0, AA=0 )
             )
demog <- setup_demography( demog, pop )
# burn-in
sim <- simulate_pop( pop, demog, times=c(0,100) )
pop$N[] <- sim$N[,,2]
```

Here is a function to initiate and track many mutations:
```{r setup_mutation_sim, cache=TRUE}
n.sims <- 20
points.per.sim <- 20
sim_estab <- function (pop, demog) {
    nhood.radius <- 1000
    many.sims <- lapply( 1:n.sims, function (dummy) {
            nhoods <- sample_neighborhoods( x=mask(habitat,habitat,maskvalue=0), n=points.per.sim, 
                                   radius=nhood.radius, separation=2*nhood.radius )
            pop <- set_N( pop, i=which(pop$habitable), j="aA", 
                     value=ifelse(which(pop$habitable)%in%nhoods$center.cells,1,0) )
            sim <- simulate_pop( pop, demog, times=c(1,50,100),
                             summaries=c( census_neighborhoods( pop, nhoods ),  
                                        list( totals=function(N){colSums(N)} ) ),
                         stop.fun=function(N){ sum(N)==0 } )
            return( list( centers=nhoods$centers, summaries=sim$summaries ) )
         } )
    sim.results <- list( centers = SpatialPoints( do.call(rbind,lapply(lapply(many.sims,"[[","centers"),coordinates)) ),
                        established = do.call( c, lapply( many.sims, function (ms)
                           {( (ms$summaries[["aA"]]+ms$summaries[["AA"]])[nrow(ms$summaries[["aA"]]),] > 0 )} ) )
                        )
    return( sim.results )
}
```

Now let's apply it across a range of $s$:
```{r different_s, cache=TRUE, depends="setup_mutation_sim"}
svals <- seq(.025,.4,length.out=25)
s.sims <- lapply( svals, function (s) {
                 demog$prob.germination$s <- s
                 sim_estab(pop,demog)
              } )
```

The overall probabilities of establishment are:
```{r mean_probs}
cbind( s=svals, p_estab=sapply( s.sims, function (x) { mean(x$established) } ) )
```

Here are the results:
```{r plot_interp_prob, fig.height=1.5*fig.dim, fig.width=1.5*fig.dim}
for (k in seq_along(s.sims)) {
    plot( habitat, main=sprintf("s=%f",svals[k]) )
    points(s.sims[[k]]$centers,pch=20,col=adjustcolor(1+s.sims[[k]]$established,0.5))
}
```




# Raw data:

Here are the probabilities, and also
coordinates of mutations, along with whether they established:
```{r write_results}
res.tab <- cbind( s=svals, 
                 p_estab=sapply( s.sims, function (x) { mean(x$established) } ),
                 n=sapply( s.sims, function (x) { length(x$established) } ) )
write_table_cache(res.tab)
```
```{r results, comment=NA}
dput( cbind( s=svals, p_estab=sapply( s.sims, function (x) { mean(x$established) } ) ) )
dput(svals)
dput( s.sims )
```


