---
title: "Speed of Expansion"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(
              fig.width=3*fig.dim,
              fig.height=fig.dim,
              fig.align='center',
              error=FALSE )
library(Matrix)
library(raster)
library(rgeos)
library(landsim)
```

Command-line for this file:
```{r call, echo=FALSE}
paste(commandArgs(),collapse=" ")
```

```{r analysis_setup, echo=FALSE}
required <- c("habitat")
usage <- " \
Before rendering this file, the following must be defined: \
    habitat : a RasterLayer \
but they can't be found.
"
if ( !all(sapply(required,exists)) ) { stop(usage) }
```

Here's the standard demography,
motivated by Joshua tree.
To change the selection coefficient we change `demog$prob.germination$s`.
```{r setup_demog, echo=FALSE}
demog <- demography(
        prob.seed = 0.2, fecundity = 100,
        prob.germination = vital( 
                 function (N, ...) {
                     out <- r0 / ( 1 + rowSums(N)/carrying.capacity )
                     return( cbind( aa=out, aA=(1+s)*out, AA=(1+s)^2*out ) )
                 },
                 r0 = 0.01,  s = 0.5 ),
        prob.survival = 0.9,
        pollen.migration = migration(
                 kern = function (x) { exp(-sqrt(x)) },
                 sigma = 300, radius = 1200, normalize = NULL ),
        seed.migration = migration(
                 kern = "gaussian", sigma = 100,
                 radius = 1200, normalize = 1 ),
        genotypes = c("aa","aA","AA")
    )
```

Here's the standard population, ditto:
```{r init_demog}
base.carrying.capacity <- 120 * prod(res(habitat))/270^2  # estimated from google maps near desert queen mine
habitable <- (!is.na(values(habitat)) & values(habitat)>0)
pop <- population( 
                  habitat = habitat,
                  # accessible = !is.na(values(habitat)),
                  accessible = habitable,
                  habitable = habitable,
                  genotypes = c("aa","aA","AA"),
                  carrying.capacity = base.carrying.capacity * values(habitat)[habitable],
                  N = cbind( aa=rpois(sum(habitable),base.carrying.capacity),
                             aA=0, AA=0 )
             )
demog <- setup_demography( demog, pop )
# burn-in
sim <- simulate_pop( pop, demog, times=c(0,100) )
pop$N[] <- sim$N[,,2]
```

Here are the **actual mean dispersal distances**:
```{r real_sigmas}
seed.sigma <- mean(dispersal_distance(demog$seed.migration$M,pop))
pollen.sigma <- mean(dispersal_distance(demog$pollen.migration$M,pop))
rbind(seed=seed.sigma, pollen=pollen.sigma)
```


This function will pick a location for a mutation, start a wave there, and measure its speed:
```{r run_setup}
sim_expansion <- function (pop, demog) {
    mutloc <- list( cell.number=sample.int(nhabitable(pop),1) )
    mutloc$center <- SpatialPoints(xyFromCell(pop$habitat,which(pop$habitable)[mutloc$cell.number]))

    ### place the single mutation, run it forwards for 100 generations, and if it hasn't established, try again.
    # summary functions
    # distance
    dists <- pointDistance( mutloc$center, xyFromCell(pop$habitat,which(pop$habitable)), lonlat=FALSE )
    # distance to the mutation center for each habitable cell
    volume_fun <- function (N) { colSums(N>0) }
    radius_fun <- function (N) { apply( N, 2, function (x) { suppressWarnings( max( dists[ x>0 ] ) ) } ) }
    summary.funs <- list( totals=function(N){colSums(N)}, volume=volume_fun, radius=radius_fun)
    # try until establishment
    for (ntries in 1:200) {
        pop <- set_N( pop, i=which(pop$habitable), j="aA", 
                 value=ifelse(1:nrow(pop$N)==mutloc$cell.number,1,0) )
        sim <- simulate_pop( pop, demog, times=seq(0,200,length.out=11),
                         summaries = summary.funs, stop.fun=function(N){ sum(N)==0 } )
        total.aA <- sim$summaries[[1]][nrow(sim$summaries[[1]]),"aA"]
        if ( total.aA>0 ) { break }
    }
    if (total.aA==0) { return( NULL ) }  # didn't work
    ## Run longer 
    sim <- extend_simulation( sim, pop, demog, times=seq(sim$t,1000,length.out=41),
                         summaries = summary.funs, stop.fun=function(N){ sum(N)==0 } )

    AA.lims <- c(1.2,0.8)*range(sim$summaries[["radius"]][,"AA"],finite=TRUE)
    AA.usethese <- which( ( sim$summaries[["radius"]][,"AA"] > AA.lims[1] ) 
                          & ( sim$summaries[["radius"]][,"AA"] < AA.lims[2] ) )
    if (length(AA.usethese)>1) {
        AA.speed <- diff( sim$summaries[["radius"]][range(AA.usethese),"AA"] ) / diff( sim$summary.times[range(AA.usethese)] )
    } else {
        AA.speed <- NA
    }

    return( list( sim=sim, pop=pop, speed=AA.speed ) )
}
```

Now let's apply it to different places and different values of $s$:
```{r different_s, cache=TRUE}
svals <- rep( c(.025, .05, 0.1, 0.2, 0.4), each=3 )
s.sims <- lapply( svals, function (s) {
                 demog$prob.germination$s <- s
                 sim_expansion(pop,demog)
              } )
```
```{r cull_different_s}
# remove those that didn't work
goodones <- ! sapply( s.sims, is.null )
svals <- svals[ goodones ]
s.sims <- s.sims[ goodones ]
```

```{r plot_different_s}
s.cols <- rainbow(1.5*length(unique(svals)))[match(svals,unique(svals))]
aA.radii <- sapply( s.sims, function (ss) { ss$sim$summaries$radius[,"aA"] } )
AA.radii <- sapply( s.sims, function (ss) { ss$sim$summaries$radius[,"AA"] } )
speed.segs <- do.call( rbind, lapply( s.sims, function (ss) {
                AA.lims <- c(1.2,0.8)*range(ss$sim$summaries[["radius"]][,"AA"],finite=TRUE)
                AA.usethese <- which( ( ss$sim$summaries[["radius"]][,"AA"] > AA.lims[1] ) 
                                      & ( ss$sim$summaries[["radius"]][,"AA"] < AA.lims[2] ) )
                if (length(AA.usethese)>1) {
                    AA.speed <- diff( ss$sim$summaries[["radius"]][range(AA.usethese),"AA"] ) / diff( ss$sim$summary.times[range(AA.usethese)] )
                    c(  ss$sim$summary.times[range(AA.usethese)], ss$sim$summaries[["radius"]][range(AA.usethese),"AA"] )
                } else {
                    c(NA,NA,NA,NA)
                }
         } ) )
colnames(speed.segs) <- c("x0","x1","y0","y1")

matplot( aA.radii, type='l', lty=1, col=s.cols )
matlines( AA.radii, lty=2, col=s.cols )
segments( x0=speed.segs[,"x0"], x1=speed.segs[,"x1"], y0=speed.segs[,"y0"], y1=speed.segs[,"y1"], lty=3, col=s.cols )

plot(sqrt(svals), sapply(s.sims, "[[", "speed"), xlab="sqrt(s)", ylab="speed", col=s.cols)
```

Here's the data:
```{r raw_data, comment=NA}
dput(svals)
sumtabs <- lapply( seq_along(svals), function (k) {
            with( list2env(s.sims[[k]]), {
                do.call( cbind, c( 
                          list(t=sim$summary.times), 
                          lapply( 1:3, function (j) { 
                                 x <- sim$summaries[[j]]; 
                                 colnames(x) <- paste(names(sim$summaries)[j],colnames(x),sep="_"); 
                                 x } ) 
                          ) )
                } )
      } )
dput(sumtabs)
```
