---
title: "Estimating the speed of expansion."
author: "Peter Ralph"
date: "`r Sys.Date()`"
---



```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(rgeos)
# library(gstat)
library(parallel)
devtools::load_all("../landsim")
```

```{r setup}
layerdir <- "../layers/Mojavset/Spatial/targets/biodiversity/binary"
layername <- "Eschscholzia_minutiflora_ssp_twisselmannii_broad_extent_avg.tif"
# habitats are listed with 0=absence, 1=presence.
habitat <- raster( file.path(layerdir,layername) )
```

Here's the raster:
```{r show_habitat, fig.width=fig.dim}
cropbox <- extent( c(
            xmin        = -2093993,
            xmax        = -1849272,
            ymin        =  1468496,
            ymax        =  1663602 ))
plot(habitat,main=layername)
# lines( cropbox )
habitat <- crop( habitat, cropbox )
```

To estimate the speed of expansion,
we'll: 

1.  Select a location to initiate a mutation, and place a mutation there,
2.  Run a simulation forwards a few generations to check for establishment,
3.  Plot total numbers and radius.

# Demography

First we need a demographic set-up.
```{r demog_setup}
carrying.capacity <- 10
pop <- population( 
                  habitat = habitat,
                  accessible = !is.na(values(habitat)),
                  habitable = (!is.na(values(habitat)) & values(habitat)>0),
                  genotypes = c("aa","aA","AA"),
                  N = cbind( aa=rpois_raster(carrying.capacity*mask(habitat,habitat,maskvalue=0),only.values=TRUE),
                             aA=0, 
                             AA=0 )
             )

germination.args <- list( 
             r0 = 0.01,  # one in ten seeds will germinate at low densities
             s = 1.5,    # multiplicative selective benefit of the A allele
             carrying.capacity = carrying.capacity, 
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1,
                                     do.M=TRUE,
                                     population=pop
                                 )
         )
germination_fun <- function (N, ...) {
    out <- germination.args$r0 / ( 1 + migrate(germination.args$competition,x=rowSums(N))/germination.args$carrying.capacity )
    return( cbind( aa=out, aA=germination.args$s*out, AA=germination.args$s^2*out ) )
}

demog <- demography(
        prob.seed = 0.2,
        fecundity = 100,
        prob.germination = germination_fun,
        prob.survival = 0.9,
        pollen.migration = migration(
                            kern = function (x) { exp(-sqrt(x)) },
                            sigma = 300,
                            radius = 1200,
                            normalize = NULL,
                            do.M=TRUE,
                            population=pop
                     ),
        seed.migration = migration(
                            kern = "gaussian",
                            sigma = 100,
                            radius = 1200,
                            normalize = 1,
                            do.M=TRUE,
                            population=pop
                     ),
        genotypes = c("aa","aA","AA"),
        mating = mating_tensor( c("aa","aA","AA") )
    )
```

# Initiate mutations

We'll choose a location for a single mutation.
```{r init_muts, fig.width=fig.dim}
mutloc <- list( cell.number=sample.int(nhabitable(pop),1) )
mutloc$center <- SpatialPoints(xyFromCell(pop$habitat,which(pop$habitable)[mutloc$cell.number]))
plot(pop$habitat)
points(mutloc$center,pch="*",col='red')
```

# Simulate, check for establishment

We'll place the single mutation, run it forwards for 100 generations, and if it hasn't established, try again.
```{r sim_muts, cache=TRUE}
for (ntries in 1:100) {
    pop <- set_N( pop, i=which(pop$habitable), j="aA", 
             value=ifelse(1:nrow(pop$N)==mutloc$cell.number,1,0) )
    sim <- simulate( pop, demog, times=seq(0,100,length.out=11),
                     summaries= list( totals=function(N){colSums(N)} ) )
    total.aA <- sim$summaries[[1]][nrow(sim$summaries[[1]]),"aA"]
    cat("Attempt ", ntries, ": total Aa is ", total.aA, "\n")
    if ( total.aA>0 ) { break }
}
```

# Run longer 

Now, we'll run it for longer.
```{r run_longer, cache=TRUE, depends="sim_muts"}
sim <- extend_simulation( sim, pop, demog, times=seq(sim$t,1000,length.out=91),
                 summaries= list( totals=function(N){colSums(N)} ) )
matplot( sim$summaries[[1]], type='l', lty=1 )
legend("topright",lty=1,col=1:3,legend=pop$genotypes)
```

```{r anim_run, fig.show="animate", aniopts="controls", interval=0.1, fig.width=3*fig.dim}
pl <- function (x,...) { hab <- do.call(stack,list(habitat)[rep(1,NCOL(x))]); values(hab)[pop$habitable] <- x; plot(hab,nr=1,...) }
for (i in seq_along(sim$times)) {
    pl( sim$N[,,i], main=paste(pop$genotypes,c("",sprintf("t=%d",floor(sim$times[i])),"")), zlim=range(sim$N) )
}
```

# Speed of expansion

To get the speed of expansion
we'll compare the spread to the minimum enclosing circle,
for heterozygotes and for homozygotes.
```{r expansion}
dists <- pointDistance( mutloc$center, xyFromCell(pop$habitat,which(pop$habitable)), lonlat=FALSE )
max.dists <- sapply( pop$genotypes[-1], function (geno) {
                    apply( sim$N[,match(geno,pop$genotypes),], 2, function (x) { if (any(x>0)) { max( dists[ x>0 ] ) } else { NA } } )
                 } )
AA.lims <- c(1.2,0.8)*range(max.dists[,"AA"],finite=TRUE)
AA.usethese <- which( ( max.dists[,"AA"] > AA.lims[1] ) & ( max.dists[,"AA"] < AA.lims[2] ) )
AA.speed <- diff( max.dists[range(AA.usethese)] ) / diff( sim$times[range(AA.usethese)] )
matplot(sim$times,max.dists,type='l',lty=1,xlab="generation")
legend("bottomright",legend=pop$genotypes[-1],lty=1,col=1:2)
abline(0,AA.speed)
AA.speed
```
The wave is moving at about `r AA.speed` meters per generation;
the *A* allele has advantage `r germination.args$s-1`;
the mean migration distance is about `r floor((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2)` meters;
so the predicted speed is `r floor((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2*sqrt(germination.args$s-1)/(1-demog$prob.survival))` meters per generation.
(Really, one generation is `r 1/(1-demog$prob.survival)` of the time steps we here call "generations".)


# Width of the wave

To compute the width of the wave,
we'll look at two measures:

1.  The difference between the furthest occurence of an *A* allele and the furthest location where *A* is at 50%;
2.  The difference between the furthest occurence that *A* is at 25% and the furthest location it is at 75%.

Here we compute these:
```{r wave_width}
A.freqs <- (sim$N[,2,]/2 + sim$N[,3,])/(sim$N[,1,] + sim$N[,2,] + sim$N[,3,])
quant.dists <- cbind(
                     max = apply( A.freqs, 2, function (x) { if (any(x[!is.na(x)]>0.5)) { max( dists[x>0], na.rm=TRUE ) - max( dists[x>0.5], na.rm=TRUE ) } else { NA } } ),
                     iqr = apply( A.freqs, 2, function (x) { if (any(x[!is.na(x)]>0.75)) { max( dists[x>0.25], na.rm=TRUE ) - max( dists[x>0.75], na.rm=TRUE ) } else { NA } } )
                 )
est.widths <- colMeans(quant.dists[AA.usethese,],na.rm=TRUE)
```

These two methods, respectively, give estimates of
`r est.widths[1]` and `r est.widths[2]` meters, respectively.
Theory predicts this width should be $\sigma/\sqrt{s}$,
which is in this case about
`r floor((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2/sqrt(germination.args$s-1))` meters.


# Comparison to theory

Simulation doesn't match up with theory so well above.
But, we're using a naive estimate of the dispersal distance, $\sigma$.
We can get a better estimate from the migration matrix itself;
```{r empirical_sigma,fig.width=2*fig.dim}
seed.sigmas <- dispersal_distance(demog$seed.migration$M,pop)
pollen.sigmas <- dispersal_distance(demog$pollen.migration$M,pop)
mean.sigma <- (mean(seed.sigmas)+mean(pollen.sigmas))/2
c(
  empirical=mean.sigma,
  naive=(demog$pollen.migration$sigma+demog$seed.migration$sigma)/2 )
layout(t(1:2))
pl(seed.sigmas,main="sigma for seeds")
pl(pollen.sigmas,main="sigma for pollen")
```
As expected, the value is `r floor(mean.sigma)` meters,
substantially larger than the naive value of `r floor((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2)` meters.
Using this value, we have
```{r compare}
s <- (germination.args$s-1)
ctab <- data.frame(
           speed = c( expected = 2 * mean.sigma * sqrt(s) / (1-demog$prob.survival),
                      observed = AA.speed ),
           width = c( expected = 2 * mean.sigma / sqrt(s) ,
                      observed = est.widths[2] ) )
floor(ctab)
```

