---
title: "Summaries, on flatspace"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---



```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(rgeos)
# library(gstat)
library(parallel)
devtools::load_all("../landsim")
```


Here is flatspace:
```{r setup_layer}
diam <- 10e3
habitat <- raster(xmn=-diam, xmx=diam, ymn=-diam, ymx=diam, 
      resolution=100,
      crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
values(habitat) <- 20
plot(habitat)
```


# Demography

First we need a demographic set-up.
```{r demog_setup}
carrying.capacity <- 10
pop <- population( 
                  habitat = habitat,
                  accessible = !is.na(values(habitat)),
                  habitable = (!is.na(values(habitat)) & values(habitat)>0),
                  genotypes = c("aa","aA","AA"),
                  N = cbind( aa=rpois_raster(carrying.capacity*mask(habitat,habitat,maskvalue=0),only.values=TRUE),
                             aA=0, 
                             AA=0 )
             )

germination_fun <- vital( 
         function (N, ...) {
             out <- r0 / ( 1 + migrate(competition,x=rowSums(N))/carrying.capacity )
             return( cbind( aa=out, aA=s*out, AA=s^2*out ) )
         },
         r0 = 0.01,  # one in ten seeds will germinate at low densities
         s = 1.5,    # multiplicative selective benefit of the A allele
         carrying.capacity = carrying.capacity, 
         competition = migration(
                                 kern="gaussian",
                                 sigma=100,
                                 radius=300,
                                 normalize=1,
                                 do.M=TRUE,
                                 population=pop
                             )
     )

demog <- demography(
        prob.seed = 0.2,
        fecundity = 100,
        prob.germination = germination_fun,
        prob.survival = 0.9,
        pollen.migration = migration(
                            kern = function (x) { exp(-sqrt(x)) },
                            sigma = 300,
                            radius = 1200,
                            normalize = NULL,
                            do.M=TRUE,
                            population=pop
                     ),
        seed.migration = migration(
                            kern = "gaussian",
                            sigma = 100,
                            radius = 1200,
                            normalize = 1,
                            do.M=TRUE,
                            population=pop
                     ),
        genotypes = c("aa","aA","AA"),
        mating = mating_tensor( c("aa","aA","AA") )
    )
```

# Speed of expansion


Choose a location for a single mutation, and make sure it establishes:
```{r init_muts, fig.width=fig.dim, cache=TRUE}
mutloc <- list( cell.number=sample.int(nhabitable(pop),1) )
mutloc$center <- SpatialPoints(xyFromCell(pop$habitat,which(pop$habitable)[mutloc$cell.number]))
for (ntries in 1:100) {
    pop <- set_N( pop, i=which(pop$habitable), j="aA", 
             value=ifelse(1:nrow(pop$N)==mutloc$cell.number,1,0) )
    sim <- simulate_pop( pop, demog, times=seq(0,100,length.out=11),
                     summaries= list( totals=function(N){colSums(N)} ) )
    total.aA <- sim$summaries[[1]][nrow(sim$summaries[[1]]),"aA"]
    cat("Attempt ", ntries, ": total Aa is ", total.aA, "\n")
    if ( total.aA>0 ) { break }
}
```

Now, run longer:
```{r run_longer, cache=TRUE}
sim <- extend_simulation( sim, pop, demog, times=seq(sim$t,1000,length.out=91),
                 summaries= list( totals=function(N){colSums(N)} ) )
```
```{r plot_run_longer, cache=TRUE, dependson="run_longer", fig.show="animate", aniopts="controls", interval=0.1, fig.width=3*fig.dim}
pl <- function (x,...) { hab <- do.call(stack,list(habitat)[rep(1,NCOL(x))]); values(hab)[pop$habitable] <- x; plot(hab,nr=1,...) }
for (i in seq_along(sim$times)) {
    pl( sim$N[,,i], main=paste(pop$genotypes,c("",sprintf("t=%d",floor(sim$times[i])),"")), zlim=range(sim$N) )
}
```

## Speed of expansion

To get the speed of expansion
we'll compare the spread to the minimum enclosing circle,
for heterozygotes and for homozygotes.
```{r expansion}
dists <- pointDistance( mutloc$center, xyFromCell(pop$habitat,which(pop$habitable)), lonlat=FALSE )
max.dists <- sapply( pop$genotypes[-1], function (geno) {
                    apply( sim$N[,match(geno,pop$genotypes),], 2, function (x) { if (any(x>0)) { max( dists[ x>0 ] ) } else { NA } } )
                 } )
AA.lims <- c(1.2,0.8)*range(max.dists[,"AA"],finite=TRUE)
AA.usethese <- which( ( max.dists[,"AA"] > AA.lims[1] ) & ( max.dists[,"AA"] < AA.lims[2] ) )
AA.speed <- diff( max.dists[range(AA.usethese)] ) / diff( sim$times[range(AA.usethese)] )
matplot(sim$times,max.dists,type='l',lty=1,xlab="generation")
legend("bottomright",legend=pop$genotypes[-1],lty=1,col=1:2)
abline(0,AA.speed)
AA.speed
```


## Width of the wave

To compute the width of the wave,
here is the difference between the furthest occurence that *A* is at 25% and the furthest location it is at 75%:
```{r wave_width}
A.freqs <- (sim$N[,2,]/2 + sim$N[,3,])/(sim$N[,1,] + sim$N[,2,] + sim$N[,3,])
quant.dists <- cbind(
                     max = apply( A.freqs, 2, function (x) { if (any(x[!is.na(x)]>0.5)) { max( dists[x>0], na.rm=TRUE ) - max( dists[x>0.5], na.rm=TRUE ) } else { NA } } ),
                     iqr = apply( A.freqs, 2, function (x) { if (any(x[!is.na(x)]>0.75)) { max( dists[x>0.25], na.rm=TRUE ) - max( dists[x>0.75], na.rm=TRUE ) } else { NA } } )
                 )
est.widths <- colMeans(quant.dists[AA.usethese,],na.rm=TRUE)
```


## Comparison to theory

**Generation time:**
We also need the generation time,
because what we're calling a generation isn't, really.
Generation time in this simulation is determined by the probability of death per generation,
which here is `r demog$prob.survival`.

**Dispersal distance:**
Simulation doesn't match up with theory so well above.
But, we're using a naive estimate of the dispersal distance, $\sigma$.
We can get a better estimate from the migration matrix itself;
```{r empirical_sigma,fig.width=2*fig.dim}
seed.sigmas <- dispersal_distance(demog$seed.migration$M,pop)
pollen.sigmas <- dispersal_distance(demog$pollen.migration$M,pop)
mean.sigma <- (mean(seed.sigmas)+mean(pollen.sigmas))/2
c(empirical=mean.sigma,naive=(demog$pollen.migration$sigma+demog$seed.migration$sigma)/2)
layout(t(1:2))
pl(seed.sigmas,main="sigma for seeds")
pl(pollen.sigmas,main="sigma for pollen")
s <- (demog$prob.germination$s-1)
ctab <- data.frame(
           speed = c( expected = 2 * mean.sigma * sqrt(s) / (1-demog$prob.survival) ,
                      observed = AA.speed ),
           width = c( expected = 2 * mean.sigma / sqrt(s) ,
                      observed = est.widths[2] ) )
floor(ctab)
```


# Probability of Establishment

## Sample mutation locations


Sample and initiate mutations:
```{r sample_muts, fig.width=fig.dim}
nhood.radius <- 5*((demog$pollen.migration$sigma+demog$seed.migration$sigma)/2*sqrt(demog$prob.germination$s-1))
many.sims <- mclapply( 1:36, function (dummy) {
        nhoods <- sample_neighborhoods( x=mask(habitat,habitat,maskvalue=0), n=60, 
                               radius=nhood.radius, separation=2*nhood.radius )
        pop <- set_N( pop, i=which(pop$habitable), j="aA", 
                 value=ifelse(which(pop$habitable)%in%nhoods$center.cells,1,0) )
        sim <- simulate_pop( pop, demog, times=c(1,50,100),
                         summaries=c( census_neighborhoods( pop, nhoods ),  
                                    list( totals=function(N){colSums(N)} ) ) )
        return( list( centers=nhoods$centers, summaries=sim$summaries ) )
     }, mc.cores=detectCores() )
sim.results <- list( centers = SpatialPoints( do.call(rbind,lapply(lapply(many.sims,"[[","centers"),coordinates)) ),
                    established = do.call( c, lapply( many.sims, function (ms)
                       {( (ms$summaries[["aA"]]+ms$summaries[["AA"]])[nrow(ms$summaries[["aA"]]),] > 0 )} ) )
                    )
```

As a result, we have a set of initial locations
and a binary variable (whether or not they established).
```{r interp_prob, fig.height=1.5*fig.dim, fig.width=1.5*fig.dim}
plot( pop$habitat )
points(sim.results$centers,pch=20,col=1+sim.results$established)
```

The overall probability of establishment is `r mean(sim.results$established)`.

Here's the raw results:
```{r results}
jsonlite::toJSON( cbind( coordinates(sim.results$centers), sim.results$established ) )
```

