---
title: "Central Mojave Habitat Shapes"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(
              fig.width=2*fig.dim,
              fig.height=fig.dim,
              fig.align='center',
              error=FALSE )
library(Matrix)
library(raster)
library(rgeos)
library(landsim)
library(colorspace)
library(rgdal)
```

# Habitat shape statistics

We'd like to display the following statistics:

1.  Volume-radius relationship: 
    $\eta(x,r) = | H \bigcap B_{x,r} |$

2.  Patch size:
    let $N(x,r)$ be the area of everything 
    reachable without taking a jump of distance greater than $r$
    
```{r ex_habitat_setup}
area.units <- 1e4  # work in hectares
vegmap <- readOGR("../layers/desert_veg/central","ds166")
source("../layers/layer_utils.R",chdir=TRUE)
counties <- get_counties(vegmap)
elev <- get_dem(vegmap)
shade <- get_elev(vegmap)
```

## Volume-radius relationship

Here's how to find the sum of the habitat values within a given circle:
```{r one_circle, fig.width=fig.dim}
vegtype <- "Mesquite"
thisveg <- subset( vegmap, LABEL_1==vegtype )
refpoints <- spsample( thisveg, n=1, type='random' )
refcircles <- make_circles( refpoints, radii=15000, proj4string=CRS(proj4string(refpoints)) )
habcircles <- gIntersection( refcircles, thisveg )
refarea <- sum( sapply( habcircles@polygons, function (x) { x@area } ) )/area.units

# for plotting
nhab <- crop( thisveg, 2*extent(refcircles) )

plot( shade, col=adjustcolor(grey(seq(0,1,length.out=101)),0.5), 
     legend=FALSE, xlim=c(xmin(nhab),xmax(nhab)), ylim=c(ymin(nhab),ymax(nhab)) )
plot( nhab, col="grey", border=NA, add=TRUE )
plot( habcircles, col=adjustcolor("red",0.5), add=TRUE, border=NA )
lines(counties,lty=2,col=adjustcolor("black",0.5))
lines(elev,col=adjustcolor('grey',0.5))
lines(refcircles, col='red')
points(refpoints,pch="*", col='red')
text( refpoints, labels=sprintf("intersection area: %0.0f ha", refarea), pos=4 )
```

Here's what it looks like for many radii:
```{r stats}
radii <- sort( 1000 * c( 0.125, 0.25, 0.5, 1, 2, 4, 8, 16 ), decreasing=TRUE ) # in km
habcircle.list <- vector(length(radii),mode='list')
refcircles <- make_circles( refpoints, radii=radii[1], proj4string=CRS(proj4string(refpoints)) )
habcircle.list[[1]] <- gIntersection( refcircles, thisveg, byid=c(TRUE,FALSE) )
for (k in seq_along(radii)[-1]) {
    refcircles <- make_circles( refpoints, radii=radii[k], proj4string=CRS(proj4string(refpoints)) )
    habcircle.list[[k]] <- ( gIntersection( refcircles, habcircle.list[[k-1]], byid=c(TRUE,FALSE) ) )
} 
refarea <- sapply( habcircle.list, function (habcircles) {
          sapply( seq_along(habcircles), function (k) {
                            sum( sapply( habcircles[k]@polygons, function (x) { x@area } ) )/area.units
              } )
    } )

layout(t(1:2))
nhab <- crop( thisveg, 2*extent(habcircle.list[[1]]) )
plot( shade, col=adjustcolor(grey(seq(0,1,length.out=101)),0.5), 
     legend=FALSE, xlim=c(xmin(nhab),xmax(nhab)), ylim=c(ymin(nhab),ymax(nhab)) )
plot( nhab, col=grey(seq(1,0.5,length.out=100)), add=TRUE )
points( refpoints, col='red', cex=2 )
for (k in seq_along(radii)) {
    plot(habcircle.list[[k]], col=adjustcolor("red",0.5), border=NA, add=TRUE)
    circles <- make_circles( refpoints, radii=radii[k], proj4string=CRS(proj4string(nhab)) )
    lines(circles, col='red')
}
lines(counties,lty=2,col=adjustcolor("black",0.5))
lines(elev,col=adjustcolor('grey',0.5))
plot( cbind( radius=radii, area=refarea ), type='b' )
lines( seq(0,max(radii),length.out=300), pi*seq(0,max(radii),length.out=300)^2/area.units, col='red' ) 
```

Now, we'll do it for many points:
```{r many_points, cache=TRUE}
refpoints <- spsample( thisveg, n=40, type='random' )  # this will take about 90 sec/ point
refcircles <- make_circles( refpoints, radii=radii[1], proj4string=CRS(proj4string(refpoints)) )
habcircles <- gIntersection( refcircles, thisveg, byid=c(TRUE,FALSE) )
refarea <- matrix( 0, nrow=length(refpoints), ncol=length(radii) )
refarea[,1] <- sapply( seq_along(habcircles), function (k) {
                sum( sapply( habcircles[k]@polygons, function (x) { x@area } ) )/area.units
  } )
for (j in seq_along(radii)[-1]) {
    refcircles <- make_circles( refpoints, radii=radii[j], proj4string=CRS(proj4string(refpoints)) )
    habcircles <- ( gIntersection( refcircles, habcircles, byid=c(TRUE,FALSE) ) )
    refarea[,j] <- sapply( seq_along(habcircles), function (k) {
                    sum( sapply( habcircles[k]@polygons, function (x) { x@area } ) )/area.units
      } )
}

# summaries
sumarea <- t( apply( refarea, 2, quantile, c(0.05, 0.25, 0.5, 0.75, 0.95) ) )

layout(t(1:2))
plot( shade, col=adjustcolor(grey(seq(0,1,length.out=101)),0.25), legend=FALSE )
plot( thisveg, col=grey(seq(.8,0.25,length.out=100)), border=NA, add=TRUE )
points( refpoints, col='red', cex=2 )
lines(counties,lty=2,col=adjustcolor("black",0.5))
lines(elev,col=adjustcolor('grey',0.5))
matplot( radii, t(refarea), type='l', xlab='radius (m)', ylab='area (ha)', lty=1 )
lines( seq(0,max(radii),length.out=300), pi*seq(0,max(radii),length.out=300)^2/area.units, lty=3, lwd=2 ) 
matlines( radii, sumarea, lty=2, lwd=2 )
```

Now, we'll do this for each vegetation type:
```{r all_habitats, cache=TRUE}
vegtype.areas <- tapply( vegmap$AREA_, vegmap$LABEL_1, sum )
vegtype.list <- names( sort(vegtype.areas,decreasing=TRUE)[1:15] )
poly.area <- function (poly) { sapply( seq_along(poly), function (k) { sum( sapply( poly[k]@polygons, function (x) { x@area } ) ) } ) }
area.results <- lapply( vegtype.list, function (vegtype) {
        thisveg <- gBuffer( subset( vegmap, LABEL_1==vegtype ), width=0 )
        refpoints <- spsample( thisveg, n=100, type='random' )
        refcircles <- make_circles( refpoints, radii=radii[1], proj4string=CRS(proj4string(refpoints)) )
        habcircles <- gIntersection( refcircles, thisveg, byid=c(TRUE,FALSE) )
        refarea <- matrix( 0, nrow=length(refpoints), ncol=length(radii) )
        refarea[,1] <- poly.area(habcircles)/area.units
        for (j in seq_along(radii)[-1]) {
            refcircles <- make_circles( refpoints, radii=radii[j], proj4string=CRS(proj4string(refpoints)) )
            habcircles <- ( gIntersection( refcircles, habcircles, byid=c(TRUE,FALSE) ) )
            refarea[,j] <- poly.area(habcircles)/area.units
        }
        return( refarea )
    } )
names( area.results ) <- vegtype.list
```

Here are the results:
```{r all_results}
area.05 <- sapply( area.results, apply, 2, quantile, 0.05 )
area.25 <- sapply( area.results, apply, 2, quantile, 0.25 )
area.50 <- sapply( area.results, apply, 2, quantile, 0.5 )
area.75 <- sapply( area.results, apply, 2, quantile, 0.75 )
area.95 <- sapply( area.results, apply, 2, quantile, 0.95 )
# plot
pp <- function (bottom,top,...) { polygon( x=c(radii,rev(radii)), y=c(bottom,rev(top)), ... ) }
lcols <- rainbow_hcl(n=length(vegtype.list))
layout(t(1:2),widths=c(3,1))
matplot( radii, area.50, lty=1, col=lcols, ylim=range(0,area.95) )
for (k in seq_along(vegtype.list)) {
    pp( area.25[,k], area.75[,k], col=adjustcolor(lcols[k],0.5) )
    pp( area.05[,k], area.95[,k], col=adjustcolor(lcols[k],0.2) )
}
opar <- par(mar=c(par("mar"),0)[c(1,5,3,5)])
plot(0,xaxt='n',yaxt='n',xlab='',ylab='')
legend("topleft", lty=1, col=lcols, legend=vegtype.list, cex=0.5)
par(opar)
```

Here is the *proportion of the circle covered* as a function of its radius.
```{r proportion_covered}
mean.areas <-  sapply( area.results, colMeans )
layout(t(1:2),width=c(3,1))
matplot(radii, area.units*mean.areas/(pi*radii^2), type='l', log='x', ylab="proportion of circle covered", xlab="radius")
opar <- par(mar=c(par("mar"),0)[c(1,5,3,5)])
plot(0,xaxt='n',yaxt='n',xlab='',ylab='',type='n')
legend("topleft",legend=colnames(mean.areas),lty=1:5,col=1:6)
```


```{r table_results, comment=NA}
dput( area.results )
```
