---
title: "Raster habitat shapes"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(
              fig.width=2*fig.dim,
              fig.height=fig.dim,
              fig.align='center',
              error=FALSE )
library(Matrix)
library(raster)
library(rgeos)
library(landsim)
library(RColorBrewer)
library(rgdal)
```

# Habitat shape statistics

We'd like to display the following statistics:

1.  Volume-radius relationship: 
    $\eta(x,r) = | H \bigcap B_{x,r} |$

2.  Patch size:
    let $N(x,r)$ be the area of everything 
    reachable without taking a jump of distance greater than $r$
    
```{r ex_habitat_setup}
area.units <- 1e4  # work in hectares
vegmap <- readOGR("../layers/desert_veg/western/ds735.gdb","ds735")
```

## Volume-radius relationship

Here's how to find the sum of the habitat values within a given circle:
```{r one_circle, fig.width=fig.dim}
vegtype <- "Madrean Warm Semi-Desert Wash Woodland/Scrub"
thisveg <- subset( vegmap, NVCSMG==vegtype )
refpoints <- spsample( thisveg, n=1, type='random' )
refcircles <- make_circles( refpoints, radii=15000, proj4string=CRS(proj4string(refpoints)) )
habcircles <- gIntersection( refcircles, thisveg )
refarea <- sum( sapply( habcircles@polygons, function (x) { x@area } ) )/area.units

# for plotting
nhab <- crop( thisveg, 2*extent(refcircles) )

plot( nhab, col="grey", border=NA )
plot( habcircles, col=adjustcolor("red",0.5), add=TRUE, border=NA )
lines(refcircles, col='red')
points(refpoints,pch="*", col='red')
text( refpoints, labels=sprintf("intersection area: %0.0f ha", refarea) )
```

Here's what it looks like for many radii:
```{r stats}
radii <- 1000 * c( 0.5, 1, 2, 4, 8, 10 )  # in km
refpoints <- sample_neighborhoods( habitat, n=1, radius=radii[length(radii)] )
refarea <- sapply( radii, function (r) {
        sapply( seq_len(length(refpoints$centers)), function (k) {
              extract( habitat, refpoints$centers[k], buffer=r, fun=sum ) } )/area.units
    } )

layout(t(1:2))
nhab <- crop( habitat, 2*extent(refpoints$neighborhoods) )
plot( nhab, col=grey(seq(1,0.5,length.out=100)), legend=FALSE )
points( refpoints$centers, pch="*" )
for (k in seq_along(radii)) {
    circles <- make_circles( refpoints$centers, radii=radii[k], proj4string=CRS(proj4string(nhab)) )
    ## offset for some reason??
    # plot( mask( nhab, circles ), col=adjustcolor(colorRampPalette(c("white","red"))(64),alpha=0.2), add=TRUE, legend=FALSE )
    lines(circles)
}
plot( cbind( radius=radii, area=refarea ), type='b' )
lines( seq(0,max(radii),length.out=300), pi*seq(0,max(radii),length.out=300)^2/area.units, col='red' ) 
```

Now, we'll do it for many points:
```{r many_points, cache=TRUE}
refpoints <- sample_neighborhoods( mask(habitat,habitat,maskvalue=0), n=100, radius=1000 )  # takes about a second per center
refarea <- sapply( radii, function (r) {
        sapply( seq_len(length(refpoints$centers)), function (k) {
              extract( habitat, refpoints$centers[k], buffer=r, fun=sum ) } )/area.units
    } )
# summaries
sumarea <- t( apply( refarea, 1, quantile, c(0.05, 0.25, 0.5, 0.75, 0.95) ) )

layout(t(1:2))
plot(habitat)
points(refpoints$centers)
lines(refpoints$neighborhoods)
matplot( radii, t(refarea), type='l', xlab='radius (m)', ylab='area (ha)', lty=1 )
lines( seq(0,max(radii),length.out=300), pi*seq(0,max(radii),length.out=300)^2/area.units, lty=3, lwd=2 ) 
matlines( sumarea, lty=2, lwd=2, col='green' )
```

Now, we'll do this for each habitat:
```{r all_habitats, cache=TRUE}
layer.files <- list.files("../layers/cleaned",full.names=TRUE)
area.results <- lapply( layer.files, function (lf) {
        habitat <- raster("../layers/cleaned/desert_veg-western-scrub.tif")
        refpoints <- sample_neighborhoods( habitat, n=100, radius=300 )  # takes about a second per center
        refarea <- sapply( radii, function (r) {
                sapply( seq_len(length(refpoints$centers)), function (k) {
                      extract( habitat, refpoints$centers[k], buffer=r, fun=sum ) } )/area.units
            } )
        return( refarea )
    } )
names( area.results=basename(layer.files) )
```

Here are the results:
```{r all_results}
area.05 <- sapply( area.results, quantile, 0.05 )
area.25 <- sapply( area.results, quantile, 0.25 )
area.50 <- sapply( area.results, quantile, 0.5 )
area.75 <- sapply( area.results, quantile, 0.75 )
area.95 <- sapply( area.results, quantile, 0.95 )
# plot
pp <- function (bottom,top,...) { polygon( x=c(radii,rev(radii)), y=c(rev(bottom),top), ... ) }
lcols <- brewer.pal(n=length(layer.files),pal="Set2")
matplot( radii, area.50, lty=1, col=lcols, ylim=range(0,area.95) )
for (k in seq_along(layer.files)) {
    pp( area.25[,k], area.75[,k], col=adjustcolor(lcols[k],0.5) )
    pp( area.05[,k], area.95[,k], col=adjustcolor(lcols[k],0.2) )
}
legend("topleft", lty=1, col=lcols, legend=basename(layer.files))
```
