---
title: "Trimming an Input Raster"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---


```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(rgeos)
library(parallel)
devtools::load_all("../landsim")
```


The problem at hand is given a raster of predicted habitat,
to pull out a sufficiently contiguous chunk,
trimming out possibly tiny, isolated pieces.
In the process, we can get statistics about how isolated different pieces are.

To do this:

1.  Pick a number of generations, $t$, and a number of individuals, $N$.
2.  Begin with mass 1 at each raster in the habitat, and mass 0 elsewhere.
3.  Migrate this mass forwards by a Geometric($1/t$) number of generations.
4.  Threshold all cells by $1/N$ and find the connected components in the result.

This is motivated by saying that two cells are in the same cluster if 
we start with $N$ individuals in one cell,
step these forwards by $t$ migration steps,
and are expected to end up with one individual in the other cell.

```{r load_raster, fig.width=fig.dim}
x <- y <- raster("Mojavset/Spatial/targets/biodiversity/binary/Yucca_brevifolia_broad_extent_avg.tif")
ngen <- 5
N <- 1000
migr <- migration( kern="gaussian", sigma=200, radius=1000, normalize=1/ngen )
for (k in 1:ceiling(log(1e-8)/log(1/ngen))) {
    y <- y + migrate_raster(y,migr) 
}
y <- (1-1/ngen)*y
values(y)[values(y)<1/N] <- NA
y.clumps <- clump( y )
plot(y.clumps)
plot(mask(x,x,maskvalue=0),add=TRUE,col='red')
```

Here are the resulting clusters:
```{r show_clusters}
z <- mask(y.clumps,x,maskvalue=0)
plot(z,col=rainbow(length(unique(values(z)))))
round(table(values(z))/sum(!is.na(values(z))),2)
```
